Precise spiking motifs in neurobiological and neuromorphic data

### 2022-11-08 manuscript.tex
Update on Overleaf.
1

% GAP
However, computational neuroscience models have suggested that the precise timing within a sequence of spikes may play a crucial role, and that neurons may be as well integrators than synchrony detectors~\citep{abeles_role_1982}. We will investigate this very hypothesis in this review. Indeed, the response of one biological neuron largely depends on the precise timing of the sequence of incoming spikes that reach its soma~\citep{paugam-moisy_computing_2012}. In comparison to a classical analogical vector of inputs, this \emph{event-based representation} observed in the neural code is essential in understanding information processing~\citep{carr_processing_1993}. For instance, it expands the capabilities of analog representations (which rely on the firing rate) by considering a representation based on repetitions of spiking motifs at precise times of occurrence. This hypothesis is directly inspired by neurobiological observations, particularly in the hippocampus, and it expands the capabilities of analog representations (which are based on the firing rate) by considering a representation based on repetitions of spiking motifs at precise times of occurrence. Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and also suggest the existence of repeated spatio-temporal spiking motifs in the spike train reaching neurons.  A mathematical formalization would be particularly well suited to neuromorphic computing~\citep{roy_towards_2019}, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Importantly, validating this hypothesis would also be crucial in our understanding of neural processes.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dynamics of vision and consequences on the neural code}%Precise timing in the neural code enables efficient vision}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Let us start with a review of the state of the art on the role of dynamics in vision. Broadly speaking, vision is the set of processes that allow us to make sense of the luminous world, and is an intensively studied field in neuroscience, in particular with respect to deciphering the neural code. In most mammals, light enters the eye to induce neural activity on the retina, which maintains a certain similitude between the topology of external visual space and its representation on the retina, called retinotopy. The origins of this neuroscientific question can be found in the first experiments from Pierre Flourens showing the relationship between visual sensations and activity in the cerebral cortex using lesions in animals~\citep{flourens_recherches_1842,pearce_marie-jean-pierre_2009}. This was also observed by Lord Adrian when recording the activity the frog's visual system~\citep{adrian_impulses_1926}. In a series of seminal studies,~\citet{hubel_receptive_1968} showed that this activity could be selective to different features, such as orientation or motion. For a large proportion of neurons, there is remarkably a monotonic relationship between the contrast of visual features and the firing frequency of neurons. However, there is no consensus to explain the multiple nonlinear mechanisms that transform the visual scene into retinotopic neural activity maps, even though these processes seem to constitute essential pieces to this puzzle~\citep{carandini_normalization_2012}.

>>>>

% GAP
However, computational neuroscience models have suggested that the precise timing within a sequence of spikes may play a crucial role, and that neurons may be as well integrators than synchrony detectors~\citep{abeles_role_1982}. We will investigate this very hypothesis in this review. Indeed, the response of one biological neuron largely depends on the precise timing of the sequence of incoming spikes that reach its soma~\citep{paugam-moisy_computing_2012}. In comparison to a classical analogical vector of inputs, this \emph{event-based representation} observed in the neural code is essential in understanding information processing~\citep{carr_processing_1993}. For instance, it expands the capabilities of analog representations (which rely on the firing rate) by considering a representation based on repetitions of spiking motifs at precise times of occurrence. This hypothesis is directly inspired by neurobiological observations and expands the capabilities of analog representations (which are based on spiking firing rates) by considering a representation based on instances of spiking motifs at precise times of occurrence. Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and further suggest the existence of precise spatio-temporal spiking motifs in the input which excites neurons.  A mathematical formalization would be particularly well suited to neuromorphic computing~\citep{roy_towards_2019}, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Crucially, validating this hypothesis would also be crucial in our understanding of neural processes. %
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dynamics of vision and consequences on the neural code}%Precise timing in the neural code enables efficient vision}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Let us start with a focus of the state-of-the-art on the role of dynamics in vision. Broadly speaking, vision is the set of processes that allow us to make sense of the luminous world, and is an intensively studied field in neuroscience, in particular with respect to deciphering the neural code. In most mammals, light enters the eye to induce neural activity on the retina, which maintains a certain similitude between the topology of external visual space and its representation on the retina, called retinotopy. The origins of this neuroscientific question can be found in the first experiments from Pierre Flourens which, using lesions in animals, demonstrated the relationship between visual sensations and activity in the cerebral cortex~\citep{flourens_recherches_1842,pearce_marie-jean-pierre_2009}. This was also observed by Lord Adrian when recording the activity the frog's visual system~\citep{adrian_impulses_1926}. In a series of seminal studies,~\citet{hubel_receptive_1968} showed that this activity could be selective to different features, such as the visual orientation or motion of elementary contours. For a large proportion of neurons, there is remarkably a monotonic relationship between the contrast of visual features and the firing frequency of neurons. However, there is no consensus to explain the multiple nonlinear mechanisms that transform the visual scene into retinotopic neural activity maps, even though these processes seem to constitute essential pieces to this puzzle~\citep{carandini_normalization_2012}.
